---
title: "Mental Health Dataset"
subtitle: ""
author: 
  - name: "Team Data Vizionaries"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "This is project-2 in the INFO-526 course by Dr Anna Leach"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: false
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false

# install.packages("tidyverse")
# install.packages("skimr")

library(tidyverse)
library(skimr)
```

## Dataset Description

```{r}
#| label: load-dataset
#| message: false
```

::: {style="text-align: justify;"}
This project uses the ‘Olympic Medals’ dataset sourced from <a href="https://www.kaggle.com/datasets/heesoo37/120-years-of-olympic-history-athletes-and-results/" target="_blank">RGriffin’s Kaggle dataset: 120 years of Olympic history: athletes and results</a>. It is a historical dataset containing bio data on athletes and medal results from Athens 1896 to Rio 2016 and spans over 120 years of Olympic history. The original owner of the dataset, <a href="https://www.randigriffin.com/">Randi H Griffin</a>, scraped the data from the <a href="www.sports-reference.com">Sports-Reference</a> website using the code linked <a href="https://github.com/rgriff23/Olympic_history/tree/master/R">here</a>.

</br>The dataset contains 15 columns. The numerical columns: id, age, height, weight and year are represented by the numerical data type in R. The categorical columns: name, sex, team, noc, games, season, city, sport, event and medal are represented by the character data type in R. The dataset has 15 variables and 271116 entries indicating that it has a dimension of 271116 x 15. The description of the columns is as follows:
:::

| **Column Name** | **Class** | **Description**                            |
|-----------------|-----------|--------------------------------------------|
| id              | double    | Unique ID for each athlete                 |
| name            | character | Name of the athlete                        |
| sex             | character | Sex of the athlete(M or F)                 |
| age             | double    | Age of the athlete                         |
| height          | double    | Height of the athlete in cm                |
| weight          | double    | Weight of the athlete in kg                |
| team            | character | Team name                                  |
| noc             | character | National Olympic Committee (3-letter code) |
| games           | character | Olympic game year and season               |
| year            | double    | Year of olympics                           |
| season          | character | Season(Winter or Summer)                   |
| city            | character | City in which olympics is hosted           |
| sport           | character | Name of the sport                          |
| event           | character | Name of the event                          |
| medal           | character | Gold, Silver, Bronze, NA                   |

: **Data Dictionary**

```{r}

olympics <- read_csv(
  'https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2024/2024-08-06/olympics.csv'
  )
skim(olympics)
```

## Reasons for choosing the dataset

::: {style="text-align: justify;"}
The Olympics stands as one of the most iconic and prestigious sporting events in human history, a stage where nations and athletes showcase their strength, determination, and skill. We chose this dataset to uncover patterns across the games, exploring how factors like country, season, year, and athletes' physical characteristics influence performance. Through this analysis, we aim to gain deeper insights into the dynamics of global competition and athletic excellence.

**Diverse Variables**: Offers a mix of numerical (like age, height, weight) and categorical (like sex, sport, medal) variables, perfect for analyzing different aspects of the games and players involved.

**Historical Depth**: Covers over a century of Olympic Games, allowing analysis of long-term trends in athlete performance and country participation.

**Distinct Questions**: Supports distinct research questions like factors influencing medal wins and changes in global participation, with minimal overlap between variables.

**Global Significance**: The Olympics is a globally recognized event, making insights from this dataset highly relevant and interesting to a wide audience.

**Visualization Potential**: Ideal for compelling storytelling, including trends over time, comparisons across countries, and analysis of athlete attributes using clear visual distinctions.
:::

## Questions

Q1. How do the number of medals won differ between Summer and Winter Olympic Games, and what are the trends over the years?

Q2. Which countries have historically won the most medals in the Olympics, and how does the distribution vary by sport?

## Analysis plan

<div style="text-align: justify;">
Question 1: The season, medal and year variables of the dataset will be used to answer this question. A line plot will be used to visualize and explain the trends in Olympic medal count across the years for each season since it clearly illustrates the patterns. In addition to the line plot, a bar plot faceted by the season variable will be used to display the distribution of medals and look out for the reasons that resulted in discontinuities.

Question 2: We plan to use the team, noc, medal and sport variables from the dataset for visualizing the distribution. A choropleth map will be used to visually represent the number of medals won by each country and a bar chart will be used to identify and showcase the top performing countries. Moreover, an animated plot(GIF) will be used to display and track the medal distributions across years for both Summer and Winter Olympics.
</div>

## Weekly Plan

<https://docs.google.com/spreadsheets/d/1PN-CLTGIFJ8vRRejRgMa1u3bbksIbxq3S-oq-_e1FFI/edit?gid=0#gid=0>

|  |  |  |  |  |  |  |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Sr No | Task Name | Status | Assignee | Due | Priority | Summary |
| 1 | Project proposal | In Progress | All | Sep-18 | High | Write project proposal along with questions to be answered. Also complete peer review of other teams |
| 2 | Review feedback and prepare presentation | Pending | All | Sep-25 | High | Based on the received feedback, make required changes in the proposal and questions. Start with presenation as well |
| 3 | Question 1 - Problem statement and code | Pending | Rony, Abhay | Oct-02 | High | Prepare detailed explanation of what Question 1 is going to be, what columns are required and what we can conclude from the question. Based on what is to be answered, start writing code. Get it reviewed from other team members and push the changes |
| 4 | Question 2 - Problem statement and code | Pending | Kapil, Nithin | Oct-02 | High | Prepare detailed expalantion of what Question 2 is going to be, what columns are required and what we can conclude from the question. Based on what is to be answered, start writing code. Get it reviewed from other team members and push the changes |
| 5 | Question 1 - Review | Pending | Kapil, Nithin | Oct-09 | Moderate | Review the code written by other 2 members and suggest changes |
| 6 | Question 2 - Review | Pending | Rony, Abhay | Oct-09 | Moderate | Review the code written by other 2 members and suggest changes |
| 7 | Final code changes and write up | Pending | All | Oct-16 | High | Make final changes in your respective codes, push them, and prepare final write up |

: **Work Schedule**