---
title: "MindScope: Unveiling Patterns in Mental Health through Data-Driven Insights"
subtitle: ""
author: 
  - name: "Team Data Vizionaries"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "Project 2 for INFO 526 course instructed by Prof. Anna Leach"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: false
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

## Dataset Description

```{r}
#| label: load-dataset
#| message: false
```

::: {style="text-align: justify;"}
This project uses the ‘Mental Health Dataset’ sourced from Kaggle. It was prepared by Mr Bhavik Jikdara by extracting the required information from `<a href="`{=html}https://ourworldindata.org/mental-health/`" target="_blank">`{=html}Our World in Data</a> . The dataset contains a variety of features related to text analysis, psychological indicators, stress, isolation and behavioral attributes making it suitable to visualize and get insights into mental health. The data covers the time period between August 2014 and February 2016.

The dataset has a total of 17 columns: 1 datetime and 16 categorical variables. The categorical columns: Gender, Country, Occupation, self_employed, family_history, treatment, Days_Indoors, Growing_Stress, Changes_Habits, Mental_Health_History, Mood_Swings, Coping_Struggles, Work_Interest, Social_Weakness, mental_health_interview, care_options and the datetime column: Timestamp are both represented by the character data type in R. The dataset has 292364 entries and 17 columns indicating it has a dimension of 292364 x 17. The description of the columns is as follows:
:::

```{r}
dataset <- read.csv("data/mental_health.csv")
colnames(dataset)
```

| **Column Name** | **Class** | **Description** |
|----|----|----|
| Timestamp | character(datetime) | The timestamp at which an observation was recorded |
| Gender | character | Gender of the person(Male or Female) |
| Country | character | Country to which the person belongs to |
| Occupation | character | The occupation of the person |
| self_employed | character | Indicates whether the person works from themselves, has their own business(Yes or No) |
| family_history | character | History of mental health issues among family members(Yes or No) |
| treatment | character | Indicates whether a person is taking treatment for mental health issues. (Yes or No) |
| Days_Indoor | character |  |
| Growing_Stress | character | Indicates whether a person's stress level is increasing over time. (Yes or No) |
| Changes_Habits | character | It refers to changes in behavior or habits that may indicate shifts in mental health. |
| Mental_Health_History | character | Person’s mental health history |
| Mood_Swings | character | Fluctuations in a person's mood |
| Coping_Struggles | character | Difficulties a person may have in coping with mental health challenges. |
| Work_Interest | character | Indicates whether interested in doing work or not |
| Social_Weakness | character | Indicates difficulties in social interactions or relationships |
| mental_health_interview | character | Indicates whether the person would bring up the issue of mental health with potential employer in an interview |
| care_options | character | Indicates whether options are available for treatment of mental health issues, such as therapy, medication, or support groups. |

## Reasons for choosing the dataset

Mental health is a topic of increasing importance. The U.S. Department of Health and Human Services refers to mental health as a person's "emotional, psychological and social well-being." A person with good mental health has a positive outlook, can accomplish daily tasks, maintain relationships, and engage in meaningful recreation. On the other hand, severe mental health issues can affect life's essential aspects, and any unexpected event can cause a cascade of adverse effects.

This dataset has values we believe to be critical in determining a person's mental state. For example, someone working in a corporate job in the United States might be happy or have a positive mental state, but someone working at the same post in Brazil might not.

Personal factors such as days spent indoors and whether or not having a support system around also make a huge difference.

## Goal of the project

By analyzing this dataset, we hope to discover correlations and patterns between various aspects of one's life. These insights can be used to develop policies and support systems promoting mental well-being.

## Questions

1.  What correlations exist between coping difficulties, social weakness, and work interest in occupation and self-employment status? Does a person's or their family's history of mental health problems link with higher levels of mood swings, coping challenges and social fragility?

2.  What is the distribution of mental health treatment-seeking across different demographics (e.g., gender, country, occupation)? Also, are there trends seen over time (based on Timestamp), such as increasing stress levels or changes in treatment-seeking behaviors?

## Analysis plan

**Question 1**

##### Key Variables:

-   **Coping_Struggles**: Indicates difficulty in coping with mental health challenges.

-   **Social_Weakness**: Indicates difficulties in social interactions or relationships.

-   **Work_Interest**: Whether the person is interested in doing work or not.

-   **Occupation**: The occupation of the person.

-   **self_employed**: Indicates if the person is self-employed or not.

-   **family_history**: History of mental health issues among family members (may also influence coping and social weakness).

-   **Mood_Swings**: Mood fluctuations may also impact coping and social interactions.

##### Steps:

1.  **Data Preprocessing**:

    -   Convert the categorical variables into factors (e.g., `Coping_Struggles`, `Social_Weakness`, `Work_Interest`, `Occupation`, `self_employed`).

    -   Handle missing values (if any). We plan to handle the missing values depending on the percentage of missing data per feature. For features with a small percentage of missing values (\<5%), we could use imputation using the mean or median. For a moderate amount of NAs (5-30%), we will try more sophisticated methods like k-nearest neighbors (KNN) imputation or regression imputation. For features with a high percentage of missing values (\>30%), we will assess their importance to the analysis and consider dropping them if they are non-essential.

2.  **Exploratory Data Analysis (EDA)**:

    -   Visualize the distribution of `Coping_Struggles`, `Social_Weakness`, `Work_Interest` against `Occupation` and `self_employed`.

    -   Examine `Coping_Struggles` and `Social_Weakness` based on family mental health history (`family_history`).

3.  **Statistical Analysis**:

    -   **Chi-Square Test**: For categorical variables, check for relationships between `Coping_Struggles`, `Social_Weakness`, and `Work_Interest` with `Occupation` and `self_employed`.

    -   **Cross-tabulation**: Use contingency tables to analyze the relationship between coping difficulties and social weakness with work interest and employment status.

4.  **Visualizations**:

    -   **Stacked Bar Plots**: Show the distribution of coping struggles, social weakness, and work interest across `Occupation` and `self_employed`.

    -   **Heatmap**: Correlation matrix for all the categorical variables (e.g., using the `dplyr` and `ggplot2` packages in R).

    -   **Bar Plots**: To show proportions of coping and social weakness for self-employed vs. non-self-employed individuals.

    -   **Violin Plots**: To show the distribution of coping and social weakness across different occupations.

5.  **Modeling**:

    -   A **logistic regression model** could be useful to assess the likelihood of coping difficulties, social weakness, or work interest, using factors like occupation, self-employment status, and family history.

**Question 2**

##### Key Variables:

-   **treatment**: Whether a person is taking treatment for mental health issues.

-   **Gender**: Gender of the person.

-   **Country**: Country the person belongs to.

-   **Occupation**: Occupation of the person.

-   **self_employed**: Whether a person is self-employed.

-   **family_history**: History of mental health issues in the family.

-   **Timestamp**: Time when the data was collected (used to track trends over time).

-   **Growing_Stress**: Whether stress levels are increasing over time.

##### Steps:

1.  **Data Preprocessing**:

    -   Convert `Timestamp` into a Date format (using `as.Date()` in R) to allow time-based analysis.

    -   Convert categorical variables into factors for analysis.

2.  **Exploratory Data Analysis (EDA)**:

    -   Analyze the distribution of `treatment` across various demographic variables (`Gender`, `Country`, `Occupation`, `self_employed`, `family_history`).

    -   Track trends in mental health treatment-seeking behavior over time by examining the `Timestamp`.

3.  **Statistical Analysis**:

    -   **Chi-Square Test** for categorical associations between `treatment` and demographic variables.

    -   **Time Series Analysis**: Look for trends over time using `Timestamp` (e.g., increasing stress, changes in treatment-seeking behavior).

4.  **Visualizations**:

    -   **Stacked Bar Plots**: Show the proportion of individuals seeking treatment across various categories (e.g., `Gender`, `Occupation`, `self_employed`, `family_history`).

    -   **Line Plots**: Show trends in treatment-seeking behavior over time, with `Timestamp` on the x-axis and percentage of people seeking treatment on the y-axis.

    -   **Heatmap**: Visualize the correlation between `treatment` and other demographic variables.

5.  **Time Series Trends**:

    -   **Autoregressive (AR) Models** could be used to predict future trends in treatment-seeking behavior, based on historical data in the `Timestamp` field.

    -   Use **Seasonal Decomposition** to break down trends in mental health treatment-seeking behavior (if time intervals show seasonality).

6.  **Modeling**:

    -   **Logistic Regression**: To assess the likelihood of treatment-seeking behavior across different demographic variables.

    -   **Time Series Models**: To predict future trends in treatment-seeking behavior over time.

## Weekly Plan

|  |  |  |  |  |  |  |
|----|----|----|----|----|----|----|
| Sr No | Task Name | Status | Assignee | Due | Priority | Summary |
| 1 | Project proposal | In Progress | Himanshu | 2024/11/13 | 1 | Prepare project proposal highlighting questions to be answered. |
| 2 | Review feedback and prepare presentation | Pending | Kapil | 2024/12/13 | 5 |  |
| 3 | Question 1 - Problem statement and code | Awaiting review | Maksim, Abhay | 2024/11/27 | 3 |  |
| 4 | Question 2 - Problem statement and code | Awaiting review | Kapil, Himanshu | 2024/11/27 | 3 |  |
| 5 | Question 1 - Review | Awaiting review | Kapil, Himanshu | 2024/11/27 | 4 |  |
| 6 | Question 2 - Review | Awaiting review | Maksim, Abhay | 2024/11/27 | 4 |  |
| 7 | Final code changes and write up | Awaiting review | Maksim, Abhay, Kapil, Himanshu | 2024/11/28 | 2 |  |
