---
title: "Mental Health Dataset"
subtitle: ""
author: 
  - name: "Team Data Vizionaries"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "This is project-2 in the INFO-526 course by Dr Anna Leach"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: false
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

## Dataset Description

```{r}
#| label: load-dataset
#| message: false
```

::: {style="text-align: justify;"}
This project uses the ‘Mental Health Dataset’ sourced from Kaggle. It was prepared by Mr Bhavik Jikdara by extracting the required information from `<a href="`{=html}https://ourworldindata.org/mental-health/`" target="_blank">`{=html}Our World in Data</a> . The dataset contains a variety of features related to text analysis, psychological indicators, stress, isolation and behavioral attributes making it suitable to visualize and get insights into mental health.

The dataset has a total of 17 columns: 1 datetime and 16 categorical variables. The categorical columns: Gender, Country, Occupation, self_employed, family_history, treatment, Days_Indoors, Growing_Stress, Changes_Habits, Mental_Health_History, Mood_Swings, Coping_Struggles, Work_Interest, Social_Weakness, mental_health_interview, care_options and the datetime column: Timestamp are both represented by the character data type in R. The dataset has 292364 entries and 17 columns indicating it has a dimension of 292364 x 17. The description of the columns is as follows:
:::

```{r}
dataset <- read.csv("data/mental_health.csv")
colnames(dataset)
```

| **Column Name** | **Class** | **Description** |
|------------------|------------------|-------------------------------------|
| Timestamp | character(datetime) | The timestamp at which an observation was recorded |
| Gender | character | Gender of the person(Male or Female) |
| Country | character | Country to which the person belongs to |
| Occupation | character | The occupation of the person |
| self_employed | character | Indicates whether the person works from themselves, has their own business(Yes or No) |
| family_history | character | History of mental health issues among family members(Yes or No) |
| treatment | character | Indicates whether a person is taking treatment for mental health issues. (Yes or No) |
| Days_Indoor | character |  |
| Growing_Stress | character | Indicates whether a person's stress level is increasing over time. (Yes or No) |
| Changes_Habits | character | It refers to changes in behavior or habits that may indicate shifts in mental health. |
| Mental_Health_History | character | Person’s mental health history |
| Mood_Swings | character | Fluctuations in a person's mood |
| Coping_Struggles | character | Difficulties a person may have in coping with mental health challenges. |
| Work_Interest | character | Indicates whether interested in doing work or not |
| Social_Weakness | character | Indicates difficulties in social interactions or relationships |
| mental_health_interview | character | Indicates whether the person would bring up the issue of mental health with potential employer in an interview |
| care_options | character | Indicates whether options are available for treatment of mental health issues, such as therapy, medication, or support groups. |

: **Data Dictionary**

## Reasons for choosing the dataset

Fill it

## Goal of the project

Fill it

## Questions

Fill it

## Analysis plan

Fill it

## Weekly Plan

|  |  |  |  |  |  |  |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Sr No | Task Name | Status | Assignee | Due | Priority | Summary |
| 1 | Project proposal |  |  |  |  |  |
| 2 | Review feedback and prepare presentation |  |  |  |  |  |
| 3 | Question 1 - Problem statement and code |  |  |  |  |  |
| 4 | Question 2 - Problem statement and code |  |  |  |  |  |
| 5 | Question 1 - Review |  |  |  |  |  |
| 6 | Question 2 - Review |  |  |  |  |  |
| 7 | Final code changes and write up |  |  |  |  |  |

: **Work Schedule**
