---
title: "Mental Health Dataset"
subtitle: ""
author: 
  - name: "Team Data Vizionaries"
    affiliations:
      - name: "College of Information Science, University of Arizona"
description: "This is project-2 in the INFO-526 course by Dr Anna Leach"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: false
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false

# install.packages("tidyverse")
# install.packages("skimr")

library(tidyverse)
library(skimr)
```

## Dataset Description

```{r}
#| label: load-dataset
#| message: false
```

::: {style="text-align: justify;"}
This project uses the ‘Mental Health Dataset’ sourced from Kaggle. It was prepared by Mr Bhavik Jikdara by extracting the required information from `<a href="`{=html}https://ourworldindata.org/mental-health/`" target="_blank">`{=html}Our World in Data</a> . The dataset contains a variety of features related to text analysis, psychological indicators, stress, isolation and behavioral attributes making it suitable to visualize and get insights into mental health.

The dataset has a total of 17 columns: 1 datetime and 16 categorical variables. The categorical columns: Gender, Country, Occupation, self_employed, family_history, treatment, Days_Indoors, Growing_Stress, Changes_Habits, Mental_Health_History, Mood_Swings, Coping_Struggles, Work_Interest, Social_Weakness, mental_health_interview, care_options and the datetime column: Timestamp are both represented by the character data type in R. The dataset has 292364 entries and 17 columns indicating it has a dimension of 292364 x 17. The description of the columns is as follows:
:::

| **Column Name** | **Class** | **Description** |
|-----------------|-----------------|--------------------------------------|
| Timestamp | character(datetime) | The timestamp at which an observation was recorded |
| Gender | character | Gender of the person(Male or Female) |
| Country | character | Country to which the person belongs to |
| Occupation | character | The occupation of the person |
| self_employed | character | Indicates whether the person works from themselves, has their own business(Yes or No) |
| family_history | character | History of mental health issues among family members(Yes or No) |
| treatment | character | Indicates whether a person is taking treatment for mental health issues. (Yes or No) |
| Days_Indoor | character |  |
| Growing_Stress | character | Indicates whether a person's stress level is increasing over time. (Yes or No) |
| Changes_Habits | character | It refers to changes in behavior or habits that may indicate shifts in mental health. |
| Mental_Health_History | character | Person’s mental health history |
| Mood_Swings | character | Fluctuations in a person's mood |
| Coping_Struggles | character | Difficulties a person may have in coping with mental health challenges. |
| Work_Interest | character | Indicates whether interested in doing work or not |
| Social_Weakness | character | Indicates difficulties in social interactions or relationships |
| mental_health_interview | character | Indicates whether the person would bring up the issue of mental health with potential employer in an interview |
| care_options | character | Indicates whether options are available for treatment of mental health issues, such as therapy, medication, or support groups. |

: **Data Dictionary**

```{r}

dataset <- read.csv("data/mental_health.csv")

colnames(dataset)
```

## Reasons for choosing the dataset

::: {style="text-align: justify;"}
The Olympics stands as one of the most iconic and prestigious sporting events in human history, a stage where nations and athletes showcase their strength, determination, and skill. We chose this dataset to uncover patterns across the games, exploring how factors like country, season, year, and athletes' physical characteristics influence performance. Through this analysis, we aim to gain deeper insights into the dynamics of global competition and athletic excellence.

**Diverse Variables**: Offers a mix of numerical (like age, height, weight) and categorical (like sex, sport, medal) variables, perfect for analyzing different aspects of the games and players involved.

**Historical Depth**: Covers over a century of Olympic Games, allowing analysis of long-term trends in athlete performance and country participation.

**Distinct Questions**: Supports distinct research questions like factors influencing medal wins and changes in global participation, with minimal overlap between variables.

**Global Significance**: The Olympics is a globally recognized event, making insights from this dataset highly relevant and interesting to a wide audience.

**Visualization Potential**: Ideal for compelling storytelling, including trends over time, comparisons across countries, and analysis of athlete attributes using clear visual distinctions.
:::

## Questions

Q1. How do the number of medals won differ between Summer and Winter Olympic Games, and what are the trends over the years?

Q2. Which countries have historically won the most medals in the Olympics, and how does the distribution vary by sport?

## Analysis plan

::: {style="text-align: justify;"}
Question 1: The season, medal and year variables of the dataset will be used to answer this question. A line plot will be used to visualize and explain the trends in Olympic medal count across the years for each season since it clearly illustrates the patterns. In addition to the line plot, a bar plot faceted by the season variable will be used to display the distribution of medals and look out for the reasons that resulted in discontinuities.

Question 2: We plan to use the team, noc, medal and sport variables from the dataset for visualizing the distribution. A choropleth map will be used to visually represent the number of medals won by each country and a bar chart will be used to identify and showcase the top performing countries. Moreover, an animated plot(GIF) will be used to display and track the medal distributions across years for both Summer and Winter Olympics.
:::

## Weekly Plan

<https://docs.google.com/spreadsheets/d/1PN-CLTGIFJ8vRRejRgMa1u3bbksIbxq3S-oq-_e1FFI/edit?gid=0#gid=0>

|  |  |  |  |  |  |  |
|-----------|-----------|-----------|-----------|-----------|-----------|-----------|
| Sr No | Task Name | Status | Assignee | Due | Priority | Summary |
| 1 | Project proposal | In Progress | All | Sep-18 | High | Write project proposal along with questions to be answered. Also complete peer review of other teams |
| 2 | Review feedback and prepare presentation | Pending | All | Sep-25 | High | Based on the received feedback, make required changes in the proposal and questions. Start with presenation as well |
| 3 | Question 1 - Problem statement and code | Pending | Rony, Abhay | Oct-02 | High | Prepare detailed explanation of what Question 1 is going to be, what columns are required and what we can conclude from the question. Based on what is to be answered, start writing code. Get it reviewed from other team members and push the changes |
| 4 | Question 2 - Problem statement and code | Pending | Kapil, Nithin | Oct-02 | High | Prepare detailed expalantion of what Question 2 is going to be, what columns are required and what we can conclude from the question. Based on what is to be answered, start writing code. Get it reviewed from other team members and push the changes |
| 5 | Question 1 - Review | Pending | Kapil, Nithin | Oct-09 | Moderate | Review the code written by other 2 members and suggest changes |
| 6 | Question 2 - Review | Pending | Rony, Abhay | Oct-09 | Moderate | Review the code written by other 2 members and suggest changes |
| 7 | Final code changes and write up | Pending | All | Oct-16 | High | Make final changes in your respective codes, push them, and prepare final write up |

: **Work Schedule**
